One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning

### 2024-05-13 _MOB_paper.Rnw
Update affiliation
Second submission J of School Psych
Second submission J of School Psych

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups of persons showing different effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity in, for example, intervention studies. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: the generalized linear mixed model (GLMM) and item response theory (IRT). In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to longitudinal data from a study on the effects of the Head Start pre-school program, to identify subgroups of families where children show comparatively larger or smaller gains in performance. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF). DIF should be investigated before using test results for group comparisons. We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

       \textbf{Marjolein Fokkema$^1$, Mirka Henninger$^2$ and Carolin Strobl$^2$}\\
       $^1$Leiden University, $^2$Universit\"at Z\"urich

@@

Model-based recursive partitioning \citep[MOB;][]{ZeilyHoth08} % simplified wording
allows for detecting subgroups of persons that differ in the parameters of a statistical model. To find the subgroups, MOB uses \textit{recursive partitioning}, a technique that became widely known through classification and regression trees \citep[CART; ][]{Breetal:1984}. In CART, the aim is to detect groups of persons, defined by specific (combinations of) covariate values, that differ in their mean on a response variable. Detecting such subgroups may be of interest in many studies in school and developmental psychology. For example, recursive partitioning methods have been used to identify adolescents at higher risk of committing crime \citep{FrityHaup19} or experiencing corporal punishment \citep{StemyHein19}, and to identify predictors of math ability \citep{DingyZhao19}.

MOB also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters of a statistical model, instead of only the mean. For example, MOB can be used to identify children that show stronger or weaker effects of an intervention \citep{ChunyAnso21}, or stronger or weaker gains in academic skills over time \citep{FokkyZeil24}. The framework of MOB is very flexible and it can be applied to a wide range of statistical models, such as %simplfied wording 
linear and logistic regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}. In this article, we will highlight two specific MOB methods that may be particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.

Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures. For example, when children are tested at several time points, the time points are nested in children; when children from different classes from different schools participate in a study, children are nested in classes, which are in turn nested in schools. In the section \hyperref[sec:glmmtree]{MOB for Subgroup Detection in Mixed-Effects Models}, we present \hyperref[sec:TutorialMixed]{Application Example 1} to illustrate how MOB can be used to detect subgroup-specific intervention effects, while taking into account the nested data structure. In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model. We thus assume that the psychological test score(s) of interest have already been validated.

In the section~\hyperref[sec:raschtree]{MOB for Detecting DIF in Measurement Models}, we go back in the research process to the point where a new psychological test has been administered to a validation sample. Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model. Validity assessment requires that we make sure that test results are comparable, for example between children of different genders. If certain items show different measurement parameters between groups, this may put certain groups at a relative disadvantage, and these items are said to violate measurement invariance or to exhibit differential item functioning (DIF). In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, the item parameters are compared between groups of persons. This can be done in a way that allows to detect DIF, while accounting for possible true group differences in ability. Traditional approaches for testing DIF require the groups to be pre-specified in order to test for DIF. In \hyperref[sec:TutorialRasch]{Application Example 2} we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.

In the following section, we first give a short introduction into the algorithm and statistical concepts behind MOB. Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}. Our aim here is to provide an informal introduction and to illustrate the relevance of these methods for studies in school psychology. We will specify the relevant works providing the formal details of MOB and the extensions presented here in the \hyperref[sec:discussion]{Discussion} section.

@@

We illustrate the idea using a simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., a reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables. To keep the example simple, we apply MOB to a very basic global model, which comprises % simplified wording
only an intercept (or mean). It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model. Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot. Obviously, a global intercept or mean would not describe all observations equally well, there is quite some unexplained variation around the sample mean of \Sexpr{round(mean(toy_data$y), digits = 2)}.

\begin{figure}[h]
\caption{Left: Univariate distribution of the response variable. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
\begin{subfigure}[][][t]{.4\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=4.5, height=7>>=
library("colorspace")
boxplot(toy_data$y, cex = .7)
#points(x = 1, y = mean(toy_data$y), pch = 19, ylim = c(1,14))
@
\end{subfigure}
\begin{subfigure}[][][b]{.7\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=6, height=5.5>>=
library("partykit")
tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
term_node_means <- round(tapply(toy_data$y, predict(tree, type = "node"), mean), digits= 2L)
@
<<eval = FALSE, echo = FALSE, fig=TRUE, width=5, height=5>>=
plot(tree, gp = gpar(cex = .7), type = "simple")
@
\end{subfigure}
\label{fig:toy}
\end{figure}

@@

We applied MOB to the observations in the left panel of Figure~\ref{fig:toy}, specifying $x_1$, $x_2$ and $x_3$ as potential partitioning variables. The resulting tree is shown in the right panel of Figure~\ref{fig:toy}. In the root node, which contains all observations (step 1), structural change tests were performed for each of the three covariates (step 2). The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected first for splitting (step 3). The $p$ value resulting from the structural change test for the first split in $x_1$ is depicted in the root node. By default, a Bonferroni correction is applied to the $p$ values, to account for the fact that the tests are performed for all of the potential partitioning variables. Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences, while the observations within the subgroup are as similar as possible. In the right daughter node, additional significant instability was detected with respect to $x_2$, a categorical covariate. Again, the cutpoint in $x_2$ was selected so that the two resulting subgroups exhibit strongest parameter differences. In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup (step 4). The same held for nodes 4 and 5, so that the third covariate $x_3$ was never selected for splitting.

The subgroup-specific distributions of the response variable are presented in the terminal nodes at the bottom of the tree. The fact that the tree in Figure~\ref{fig:toy} displays more than one terminal node confirms that one global model for all observations does not appropriately capture the pattern in the data.  The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting. This automatic variable selection is an important characteristic of the CART and MOB algorithms. The means (intercepts) in the terminal nodes are \Sexpr{term_node_means[1]}, \Sexpr{term_node_means[2]} and \Sexpr{term_node_means[3]}, respectively. While the means and distributions in nodes 2 and 4 appear very similar, the mean in node 5 is obviously higher. Substantively, the tree suggests and interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa). If $x_1$ were age and $x_2$ gender, this would indicate that gender differences only start to occur at a later age.

@@

There are three further important characteristics of tree and MOB algorithms that we would like to mention here: The first characteristic is the type of variable and cutpoint selection an algorithm employs. Traditional classification and regression tree algorithms, like those of \citet{Breetal:1984} and \citet{Qui:1993}, performed variable and cutpoint selection in one step, which leads to an undesirable behavior called variable selection bias. That is, the traditional algorithms prefer variables offering more possible cutpoints in the selection process -- regardless of their true information content. An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}. More modern algorithms for classification and regression trees have solved this problem and offer unbiased variable selection, such as \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, which is available in \texttt{R} in packages \texttt{party} and \texttt{partykit} \citep{partykit:pkg}. The latter forms the basis for all MOB approaches presented in this paper. For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.

The second characteristic relates to the way a tree or MOB algorithm stops splitting: Modern algorithms for classification and regression trees and MOB use a criterion of statistical significance to stop splitting. Once there are no more covariates that show a significant structural change in any node, splitting is halted. In this way, the  algorithm selects only those partitioning variables that are relevant for distinguishing the groups (i.e., it performs automatic variable selection, as illustrated in the right panel of Figure~\ref{fig:toy}). Moreover, the trees will not grow as large as possible, but splitting is stopped when no more significant structural change is detected. While traditional tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (so called \textit{pruning}), modern tree and MOB algorithms employ significance tests as stopping criteria (and some can also use effect size measures, as we will see in \hyperref[sec:TutorialRasch]{Application Example 2}). This allows for stopping tree growing as soon as no significant differences can be detected anymore. Other stopping criteria are based on the number of persons in the terminal nodes. These criteria ensure that the sample sizes in the terminal nodes are large enough to estimate the statistical model in each terminal node. For reasons of interpretability, users may sometimes specify even higher requirements for the sample sizes of terminal nodes. Note that it is very well possible that no split will be implemented, when none of the specified partitioning variables show significant structural change.

The third important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner. This is a key feature of the MOB approach that makes it very flexible and sets it apart from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered. While there are phases in psychological and educational research where it is very important to specify hypotheses a-priori and test them in a confirmatory manner, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers. Still, an important challenge for the researcher remains: To specify the parametric model of interest, to specify the set of possible partitioning variables and to choose the settings of the MOB algorithm. The current paper aims to provide guidance.

@@

\section{MOB for Subgroup Detection in Mixed-Effects Models}
\label{glmmtree}

@@

GLMM trees combine MOB and generalized linear mixed-effects models (GLMMs) and were introduced by \cite{FokkySmit18}. While the 'standard' MOB trees \citep{ZeilyHoth08} allow for subgroup detection in fixed-effects GLMs, GLMM trees additionally estimate and account for random effects and can thus be used for partitioning mixed-effects regression models. Because researchers' interests in multilevel models commonly focus on the fixed effects (of time or treatment, for example), the GLMM tree algorithm only targets structural change in the fixed-effects parameters. The random-effects parameters can be specified as usual and are assumed constant; that is, they are estimated using all observations in the dataset. Structural change is thus assessed with respect to the fixed-effects parameters, while dependency between observations within the same units is accounted for by the random effects. The resulting subgroups will differ in their estimates for the fixed-effects coefficients only. As such, GLMM trees allow for detecting subgroups in multilevel models that differ with respect to any (set of) fixed-effects parameters of interest. For example, users may be interested in detecting subgroups with different means \citep{FokkyEdbr21}, but also differential effects of treatment \citep{FokkySmit18} or differential growth over time \citep{FokkyZeil24}, to name but a few examples.

The \texttt{R} functions for fitting GLMM trees are the \texttt{lmertree} (for linear mixed models) and \texttt{glmertree} (for functions from package \texttt{glmertree}. The usage of this function will be illustrated in \hyperref[sec:TutorialMixed]{Application Example 1}. Further mathematical and computational details about the GLMM tree model are also described in \cite{FokkySmit18} and \cite{FokkyZeil24}.

@@

To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start. Head Start is a federally funded nationwide pre-school program for children from low-income families in the United States. Participation in Head Start takes place from ages 3 through 5. \cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth \citep{NLSY}.

The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other pre-school program. Data from children in those families who participated in another pre-school program were excluded. The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start. The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test \citep[PPVT]{DunnyDunn81}, we will thus model PPVT trajectories over time. On average, there were \Sexpr{round(mean(table(HS_dat$ChildID)), digits = 2)} PPVT scores per child, for \Sexpr{round(100*(1-mean(table(HS_dat$ChildID) > 1)), digits = 0)}\% of the children there was only a single PPVT score available.

@@

Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's height (computed as feet x 100 + inches). Note that the latter variable is one that should be completely irrelevant for predicting performance on a vocabulary test; it is included here to illustrate that the GLMM tree algorithm can fruitfully distinguish signal from noise variables. The dataset analyzed here only comprises data from families and children for whom complete data was available.

@@

We inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None (black) versus Head Start (red). To show the global effect of age and Head Start participation, we first fit a mixed-effects model comprising their main and interaction effects, using package \texttt{lme4}. This would be the model fitted in the root node of the GLMM tree. To account for the correlation between repeated assessments on the same child, and between siblings with the same mother, we specify a random intercept for children, nested within mothers:

@@

@@

Next, we test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:

@@

With the first argument, we specified the model \verb|formula|, which has three parts separated by vertical bars: The left part (\verb|PPVT ~ Program*Age|) specifies the response variable, followed by a tilde (\verb|~|) and the fixed-effects predictors of relevance. The middle part (\texttt{1|MotherID/ChildID}) specified the random effect: Repeated PPVT assessments are nested within children, which are nested within mothers. The right part (\verb|AFTQ + Race + Income + Mom_edu_yrs + Mom_height|) specified the partitioning variables: covariates that may possibly affect the values of the fixed-effects parameters.

>>>>

\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups of persons showing different effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity in, for example, intervention studies. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: the generalized linear mixed model (GLMM) and item response theory (IRT). In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to longitudinal data from a study on the effects of the Head Start pre-school program, to identify subgroups of families where children show comparatively larger or smaller gains in performance. In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF). DIF should be investigated before using test results for group comparisons. We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on DIF effect sizes.\\}

@@

       \textbf{Marjolein Fokkema$^1$, Mirka Henninger$^{2,3}$ and Carolin Strobl$^2$}\\
       $^1$Leiden University, $^2$University of Zurich, $^3$ University of Basel

@@

Model-based recursive partitioning \citep[MOB;][]{ZeilyHoth08} % simplified wording
allows for detecting subgroups of persons that differ in the parameters of a statistical model. To find the subgroups, MOB uses \textit{recursive partitioning}, a technique that was first popularized by the classification and regression trees \citep[CART; ][]{Breetal:1984} algorithm. In CART, the aim is to detect groups of persons, defined by specific (combinations of) covariate values, that differ in their mean on a response variable. Detecting such subgroups may be of interest in many studies in school and developmental psychology. For example, recursive partitioning methods have been used to identify adolescents at risk of committing crime \citep{FrityHaup19} and experiencing corporal punishment \citep{StemyHein19}, and to identify predictors of math ability \citep{DingyZhao19}.

Like CART, MOB also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters, instead of only the mean. For example, MOB can be used to identify children that show stronger or weaker effects of an intervention \citep{ChunyAnso21}, or stronger or weaker gains in academic skills over time \citep{FokkyZeil24}. The framework of MOB is very flexible and can be applied to a wide range of statistical models, such as %simplified wording 
linear and logistic regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}. In this article, we will highlight two specific MOB methods that may be particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StroyKopf15, KomStrZei:2017:EaPM, HennyDeba23}.

Mixed-effects models become relevant whenever data are collected in longitudinal or nested data structures. For example, when children are tested at several time points, observations at each timepoint are nested in children; when children from different classes from different schools participate in a study, children are nested in classes, which are in turn nested in schools. In the section \hyperref[sec:glmmtree]{MOB for Subgroup Detection in Mixed-Effects Models}, we present an \hyperref[sec:TutorialMixed]{example application} to illustrate how MOB can be used to detect subgroup-specific intervention effects, while taking into account the nested data structure. In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model. We thus assume that the psychological test score(s) of interest have already been validated.

In the section~\hyperref[sec:raschtree]{MOB for Detecting DIF in Measurement Models}, we go back in the research process to the point where a new psychological test has been administered to a validation sample. Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model. Validity assessment requires that we make sure that test results are comparable, for example between children of different genders. If certain items show different measurement parameters between groups, this may put certain groups at a relative disadvantage. These items are said to violate measurement invariance or exhibit differential item functioning (DIF). In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, the item parameters are compared between groups of persons. This can be done in a way that allows to detect DIF, while accounting for possible true group differences in ability. Traditional approaches for testing DIF require the groups to be pre-specified in order to test for DIF. In an \hyperref[sec:TutorialRasch]{application example} we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.

In the following section, we first give a short introduction into the algorithm and statistical concepts behind MOB. Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}. Our aim here is to provide an informal introduction and to illustrate the relevance of these methods for studies in school psychology. For readers interested in the more formal details of MOB and the extensions presented here, relevant references are provided throughout the manuscript.

@@

We illustrate this idea using a simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., a reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables. To keep the example simple, we apply MOB to a very basic global model, which comprises % simplified wording
only an intercept (or mean). It would also be possible to use, for example, a linear or logistic regression model as the global model. Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot. Obviously, a global intercept or mean would not describe all observations equally well, because there is quite some unexplained variation around the sample mean of \Sexpr{round(mean(toy_data$y), digits = 2)}.

\begin{figure}[h]
\caption{Left: Univariate distribution of the response variable. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
\begin{subfigure}[][][t]{.4\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=4.5, height=7>>=
library("colorspace")
boxplot(toy_data$y, cex = .7)
@
\end{subfigure}
\begin{subfigure}[][][b]{.7\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=6, height=5.5>>=
library("partykit")
tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
term_node_means <- round(tapply(toy_data$y, predict(tree, type = "node"), mean), digits= 2L)
@
<<eval = FALSE, echo = FALSE, fig=TRUE, width=5, height=5>>=
plot(tree, gp = gpar(cex = .7), type = "simple")
@
\end{subfigure}
\label{fig:toy}
\end{figure}

\begin{figure}[h]
\caption{Top row: Residuals ordered by the values of the partitioning variables. Bottom row: Summed residuals (cumulative sums for continuous predictors, sums by level for categorical predictors).}
\begin{subfigure}[][][t]{1.2\textwidth}
<<eval = TRUE, echo = FALSE, fig=TRUE, width=12, height=4>>=
library("strucchange")
gefp_x1 <- gefp(y ~ 1, fit = lm, data = toy_data, order.by = ~ x1)
gefp_x3 <- gefp(y ~ 1, fit = lm, data = toy_data, order.by = ~ x3)
lmod <- lm(y ~ 1, data = toy_data)
resids <- residuals(lmod)
par(mfrow = c(1, 3))
plot(toy_data$x1, resids, xlab = "x1", ylab = "residuals", 
     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
plot(toy_data$x2, resids, xlab = "x2", ylab = "residuals", 
     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
plot(toy_data$x3, resids, xlab = "x3", ylab = "residuals", 
     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
@

<<eval = TRUE, echo = FALSE, fig=TRUE, width=12, height=4>>=
par(mfrow = c(1, 3))
plot(toy_data$x1[order(toy_data$x1)], cumsum(resids[order(toy_data$x1)]), 
     xlab = "x1", ylab = "cumulative sum", type = "l",
     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
boxplot(t(replicate(100, tapply(resids, toy_data$x2, sum))), 
        ylab = "sum", cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
plot(toy_data$x3[order(toy_data$x3)], cumsum(resids[order(toy_data$x3)]), 
     xlab = "x3", ylab = "cumulative sum", type = "l",
     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
@
\end{subfigure}
\label{fig:efp}
\end{figure}

@@

We applied MOB to the observations in the left panel of Figure~\ref{fig:toy}, specifying $x_1$, $x_2$ and $x_3$ as potential partitioning variables. In Step 1, the model parameters (here: the intercept) are estimated in the root node, which contains all observations. %changed
In Step 2, the model's \emph{scores} are computed. In case of an intercept-only model, these are simply the residuals. The scores are ordered with respect to the possible partitioning variables, as shown in the top row of Figure~\ref{fig:efp}. The scores should randomly fluctuate around their mean of zero, but the top row of Figure~\ref{fig:efp} suggests systematic deviations: there seems to be an association with $x_1$ and $x_2$, and little or no association with $x_3$.

MOB captures systematic deviations in the scores (i.e., structural change) by computing cumulative sums. The cumulative sums are depicted in the bottom row of Figure~\ref{fig:efp}. To decide in Step 3 whether to split the node, and if so, which variable to split on, a test statistic is computed based on the cumulative sums of each partitioning variable; full details on the computation are provided in \citep{ZeilyHoth08}. In the current example, the structural change tests revealed that $x_1$ was most strongly associated with instabilities in the intercept and it was thus selected for splitting in Step 3.

In Step 4, the procedure is repeated in each of the resulting subgroups. The resulting tree is shown in the right panel of Figure~\ref{fig:toy}. In the right subgroup (node 3), additional significant instability was detected with respect to $x_2$, a categorical covariate. Again, the cutpoint for $x_2$ was selected so that the two resulting subgroups exhibit an as large as possible parameter difference. In the left subgroup (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup. The same held for nodes 4 and 5, so that the third covariate $x_3$ was never selected for splitting.

The $p$ values resulting from the structural change tests of each split are depicted in the corresponding nodes. By default, the $p$ values are Bonferroni corrected, to account for the fact that the tests are performed for all potential partitioning variables. After selecting $x_1$ for splitting, the cutpoint is selected so that the two resulting subgroups exhibit the strongest parameter differences, while the observations within the subgroups are as similar as possible.

The subgroup-specific distributions of the response variable are depicted in the terminal nodes at the bottom of the tree. The fact that the tree in Figure~\ref{fig:toy} displays more than one terminal node confirms that one global model for all observations does not appropriately capture the patterns in the data.  The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were selected for splitting. This automatic variable selection is an important characteristic of the CART and MOB algorithms.

The means (intercepts) in the terminal nodes are \Sexpr{term_node_means[1]}, \Sexpr{term_node_means[2]} and \Sexpr{term_node_means[3]}, respectively. While the means and distributions in nodes 2 and 4 appear very similar, the mean in node 5 is obviously higher. Substantively, this tree suggests an interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa). If $x_1$ were age and $x_2$ gender, this would indicate that gender differences only start to occur at a later age.

@@

There are three further important characteristics of tree and MOB algorithms that we would like to mention here: The first characteristic is the type of variable and cutpoint selection an algorithm employs. Traditional classification and regression tree algorithms, like those of \citet{Breetal:1984} and \citet{Qui:1993}, performed variable and cutpoint selection in one step, which leads to an undesirable behavior called variable selection bias. That is, the traditional algorithms prefer variables offering more possible cutpoints in the selection process -- regardless of their true information content. An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}. More modern algorithms for classification and regression trees have solved this problem and offer unbiased variable selection, such as \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, which is available in \texttt{R} in packages \texttt{party} and \texttt{partykit} \citep{partykit:pkg}. The latter forms the basis for all MOB approaches presented in this paper. For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006}, \citet{StrMalTut:2009:PM} and /citet{StroyKopf15}.

The second characteristic relates to the way a tree or MOB algorithm stops splitting: Modern algorithms for classification and regression trees and MOB use a criterion of statistical significance to stop splitting. Once there are no more covariates that show a significant structural change in any node, splitting is halted. In this way, the  algorithm selects only those partitioning variables that are relevant for distinguishing the groups (i.e., it performs automatic variable selection, as illustrated in the right panel of Figure~\ref{fig:toy}). Moreover, the trees will not grow as large as possible, but splitting is stopped when no more significant structural change is detected. While traditional tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (so called \textit{pruning}), modern tree and MOB algorithms employ significance tests as stopping criteria (and some can also use effect size measures, as we will see in \hyperref[sec:TutorialRasch]{Application Example 2}). This allows for stopping tree growing as soon as no significant differences can be detected anymore. Other stopping criteria are based on the number of persons in the terminal nodes. These criteria ensure that the sample sizes in the terminal nodes are large enough to estimate the statistical model in each terminal node. For reasons of interpretability, users may specify even higher requirements for the sample sizes of terminal nodes.

The third characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner. This is a key feature of the MOB approach that makes it very flexible and sets it apart from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered. While there are phases in psychological and educational research where it is very important to specify hypotheses a-priori and test them in a confirmatory manner, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers. Still, an important challenge for the researcher remains: To specify the parametric model of interest, to specify the set of possible partitioning variables and to choose the settings of the MOB algorithm. The current paper aims to provide guidance.

@@

\section{MOB for Subgroup Detection in Mixed-Effects Models}
\label{sec:glmmtree}

@@

GLMM trees combine MOB and generalized linear mixed-effects models (GLMMs) and were introduced by \cite{FokkySmit18}. While the 'standard' MOB trees \citep{ZeilyHoth08} allow for subgroup detection in fixed-effects GLMs, GLMM trees additionally estimate and account for random effects and can thus be used for subgroup detection in mixed-effects regression models. The GLMM tree algorithm only targets structural change in the fixed-effects parameters. The random-effects parameters are specified as usual and assumed constant; that is, they are estimated using all observations in the dataset. The resulting subgroups will differ in their estimates for the fixed-effects coefficients only. As such, GLMM trees allow for detecting subgroups in multilevel models that differ with respect to any (set of) fixed-effects parameters of interest. For example, users may be interested in detecting subgroups with different means \citep{FokkyEdbr21}, but also differential effects of treatment \citep{FokkySmit18} or differential growth over time \citep{FokkyZeil24}, to name but a few examples.

The \texttt{R} functions for fitting GLMM trees are \texttt{lmertree} and \texttt{glmertree}. The former assumes a normally distributed response, while the latter supports, for example, binomial or count responses. The following section will illustrate the usage of \texttt{lmertree}. Mathematical and computational details of the GLMM tree algorithm are described in \cite{FokkySmit18} and \cite{FokkyZeil24}.

@@

To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in the Head Start program. Head Start is a federally funded nationwide pre-school program for children from low-income families in the United States. Participation in Head Start takes place from ages 3 through 5. \cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the 1979 cohort of the National Longitudinal Survey of Youth \citep{NLSY}.

The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other pre-school program. Data from children who participated in another pre-school program were excluded. The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start. The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test \citep[PPVT; ][]{DunnyDunn81}, for which we model trajectories over time. On average, there were \Sexpr{round(mean(table(HS_dat$ChildID)), digits = 2)} PPVT scores per child, for \Sexpr{round(100*(mean(table(HS_dat$ChildID) > 1)), digits = 0)}\% of the children there were $>1$ PPVT scores available.

@@

Our dataset contains five family characteristics: Mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; family income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's height (computed as feet x 100 + inches). Note that the latter variable is one that should be completely irrelevant for predicting performance on a vocabulary test; it is included here to illustrate that the GLMM tree algorithm can fruitfully distinguish signal from noise variables. The dataset analyzed here only comprises data from families and children for whom complete data was available.

@@

We first inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None versus Head Start. To show the global effect of age and Head Start participation, we first fit a mixed-effects model comprising their main and interaction effects, using package \texttt{lme4}. This would also be the model fitted in the root node of the GLMM tree:

@@

We specified the fixed main and interaction effects of `Program` and `Age` as `Program*Age`. The random effects were specified as `(1|MotherID/ChildID)`. The \texttt{1} specifies that a random intercept should be estimated and \texttt{MotherID/ChildID} specifies that every child and every mother should obtain a random intercept, so as to account for correlations between repeated assessments of the same child, and correlations between children of the same mother. The forward slash \texttt{/} indicates that the children are 'nested' within mothers; that is, that there are multiple children per mother, but only one mother per child.

@@

We now test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:

@@

With the first argument, we specified the model \verb|formula|, which has three parts separated by vertical bars: The left part (\verb|PPVT ~ Program*Age|) specifies the response variable, followed by a tilde (\verb|~|) and the fixed-effects predictors of relevance. The middle part (\texttt{1|MotherID/ChildID}) specified the random effects, which are identical to the random effect in the earlier mixed-effects model. The right part (\verb|AFTQ + Race + Income + Mom_edu_yrs + Mom_height|) specified the partitioning variables: covariates that may possibly affect the values of the fixed-effects parameters.

With the second argument, we specified the dataset which contain the variables. With the \texttt{cluster} argument, we account for the level at which the partitioning variables ae measured. In this example, the partitioning variables are measured at the mother (or family) level, which we indicated by specifying \texttt{cluster = MotherID}. As a result, the parameter stability tests will employ so-called clustered covariances \citep{ZeilyKoll20}. Not specifying the \texttt{cluster} argument would result in use of the default observation-level covariances in the parameter stability tests, inflating type-I error rates when partitioning variables are measured on a higher level \citep{FokkyZeil24}.
